---
title: "Assignment 3- Task 3 Santa Barbara Urchins"
author: "Sydney Rilum"
date: "10/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(kableExtra)
library(lubridate)
```

## Summary

In the document, I will create exploratory graphs and calculate measures of central tendency in order to compare the sizes of purple and red urchins. In addition, I will explore purple urchin counts by site and year.

### Read in the data

Here, I read in the urchins data set.

```{r}
urchins <- read.csv(here("data", "urchins.csv"))
```

## Section 1: Size comparisons, purple and red urchins

### 1A: Exploratory data visualization

Here, I create two exploratory graphs comparing the distribution sizes between purple and red urchins.

**Faceted histogram**
```{r}
ggplot(data = urchins, aes(x = SIZE)) +
  geom_histogram(bins = 20) +
  facet_wrap(~COMMON_NAME) +
  labs(x = "Urchin diameter (cm)", 
       y = "Number of urchins")
```

**Boxplot**
```{r}
ggplot(data = urchins, aes(x = COMMON_NAME, y = SIZE)) +
  geom_boxplot() +
  labs(y = "Urchin diameter (cm)") +
  theme(axis.title.x = element_blank())
```

### 1B: Descriptive statistics table

Here, I create a summary table that contains the mean, median, standard deviation, and sample size for purple and red urchins size.

```{r}
urchins_stats <- urchins %>% 
  group_by(COMMON_NAME) %>% 
  summarize(mean = mean(SIZE),
            median = median(SIZE),
            sd = sd(SIZE),
            n = n()) 

kbl(urchins_stats) %>% 
  kable_styling()
```

### 1C: Analysis

From the histogram, both the purple and red urchin size data appear to be normally distributed, with purple urchin sizes having a slightly narrower spread of data than red urchin size data. The boxplot indicates that the red urchin size data set has seven outliers, which could explain the wider spread of data shown in the histogram. The statistical calculations indicate that purple urchins have a smaller mean diameter of ~4.5 cm compared to red urchins, which have a mean diameter of ~6.8 cm. Both median values are close to their respective mean values, further suggesting that the both data sets are close to being symmetrically distributed. The SD of red urchin size data is slightly larger than the SD of purple urchin size, which also supports that the red urchin data has a slightly wider spread. The difference in mean urchin diameter between the species is ~2.3 cm and the percent difference is ~41%. 


## Section 2: Explore purple urchin counts by site and year

### 2A: Create a new subset of data

Here, I create a subset of urchins data that:
- Only includes observations for purple urchins
- Converts the date to class “Date” 
- Creates a new column that only contains the year
- Creates a new column that contains the full name (not abbreviations) for each of the five sites.

```{r}
purple_urchins <- urchins %>% 
  filter(COMMON_NAME == "Purple Urchin") %>% 
  mutate(DATE = mdy(DATE), .keep = "unused") %>% 
  mutate(YEAR = year(DATE)) %>% 
  mutate(site_full = case_when(
    SITE == "NAPL" ~ "Naples Reef", 
    SITE == "AQUE" ~ "Arroyo Quemado Reef",
    SITE == "CARP" ~ "Carpinteria Reef",
    SITE == "MOHK" ~ "Mohawk Reef",
    SITE == "IVEE" ~ "Isla Vista Reef"))
```

### 2B: Descriptive statistics table

Here, I make a summary table containing the total number of purple urchins observed by site and year. 

```{r}
purple_urchins_stats <- purple_urchins %>% 
  count(site_full, YEAR)

kbl(purple_urchins_stats) %>% 
  kable_styling()
```

### 2C: Exploratory graph

Here, I make an exploratory graph of counts of purple urchins recorded at each site over time for the years included in the study.

```{r}
ggplot(data = purple_urchins_stats, aes(x = YEAR, y= n, fill = site_full)) +
  geom_col() +
  facet_wrap(~site_full) +
  labs(x = "year", 
       y = "purple urchin count", 
       title = "Purple urchins observed by site and year") +
  theme_minimal() +
  theme(legend.position = "none")
```

