---
title: "Assignment 3- Task 3 Santa Barbara Urchins"
author: "Sydney Rilum"
date: "10/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(kableExtra)
library(lubridate)
```

### Read in the data

Here, I read in the urchins data set.

```{r}
urchins <- read.csv(here("data", "urchins.csv"))
```

## Section 1: Size comparisons, purple and red urchins

### 1A: Exploratory data visualization

Here, I create two graphs comparing the distribution sizes between purple and red urchins.

Faceted histogram
```{r}
ggplot(data = urchins, aes(x = SIZE)) +
  geom_histogram(bins = 20) +
  facet_wrap(~COMMON_NAME) +
  labs(x = "Urchin diameter (cm)", y = "Number of urchins")
```

Boxplot
```{r}
ggplot(data = urchins, aes(x = COMMON_NAME, y = SIZE)) +
  geom_boxplot() +
  labs(x = "Urchin type", y = "Urchin diameter (cm)")
```

### 1B: Descriptive statistics table

Here, I create a summary table that contains the mean, median, standard deviation, and sample size for purple and red urchins size.

```{r}
urchins_stats <- urchins %>% 
  group_by(COMMON_NAME) %>% 
  summarize(mean = mean(SIZE),
            median = median(SIZE),
            sd = sd(SIZE),
            n = n()) 

kable(urchins_stats) %>% 
  kable_styling()
```

### 1C: Analysis
***EDIT
Looking at the faceted histogram, the spread of the red urchin diameter values appears to be larger, while the purple urchin values are more grouped/narrower spread. Both datasets appear to be normally distributed from the histogram, but the boxplot suggests slight positive skew. This potentially indicates that red urchin diameter has a larger range of values than for purple urchins. The boxplot indicates that there are a large amount of outliers for the red urchins dataset, explaining the larger data spread.

The boxplot displays the red urchin median (~7 cm) to be larger than the purple uchin median (~5 cm). In addition, the red urchins have diameter obsrevations that go up to 12 cm +, while purple urchin diameter maxes out at 9 cm.

## Section 2: Explore purple urchin counts by site and year

### 2A: Create a new subset of data
Here, I create a subset of urchins data that:
- Only includes observations for purple urchins
- Converts the date to class “Date” 
- Creates a new column that only contains the year
- Creates a new column that contains the full name (not abbreviations) for each of the five sites.

```{r}
purple_urchins <- urchins %>% 
  filter(COMMON_NAME == "Purple Urchin") %>% 
  mutate(DATE = mdy(DATE), .keep = "unused") %>% 
  mutate(YEAR = year(DATE)) %>% 
  mutate(site_full = case_when(
    SITE == "NAPL" ~ "Naples Reef", 
    SITE == "AQUE" ~ "Arroyo Quemado Reef",
    SITE == "CARP" ~ "Carpinteria Reef",
    SITE == "MOHK" ~ "Mohawk Reef",
    SITE == "IVEE" ~ "Isla Vista Reef"))
```

### 2B: Descriptive statistics table

Here, I make a summary table containing the total number of purple urchins observed by site and year. 

```{r}
purple_urchins_stats <- purple_urchins %>% 
  count(site_full, YEAR)

purple_urchins_stats
```

### 2C: Exploratory graph

Here, I make an exploratory graph of counts of purple urchins recorded at each site over time for the years included in the study.

```{r}
ggplot(data = purple_urchins_stats, aes(x = YEAR, y= n)) +
  geom_col() +
  facet_wrap(~site_full) +
  theme(legend.position = "none") +
  labs(x = "year", y = "purple urchin count")
```

